Informing policy via dynamic models: Eliminating cholera in Haiti

### 2022-09-26 ms.Rnw
Started Clean-up of saved files

# Check if RUN_LEVEL == 3 results exists, if so, use them.
# Include a flag to check if RUN_LEVEL == 3. If so, we might want to recompute.
# TODO: Only save parameters and likelihoods, not entire MIF objects.
if (RUN_LEVEL != 3 & file.exists("model1/run_level_3/lee1_epi_fit.rds")) {
  lee1_epi_mif <- readRDS("model1/run_level_3/lee1_epi_fit.rds")
} else {
  lee1_epi_mif <- bake(
    file = paste0("model1/", rl_dir, "lee1_epi_fit.rds"),
    {
      foreach(
        i = 1:nrow(epi_params),
        .packages = c('pomp'),
        .combine = c
      ) %dopar% {
        r_params <- unlist(epi_params[i, ])
        coef(h1_epi) <- r_params
        mif2(
          h1_epi,
          Np = NP_MIF,
          Nmif = NMIF,
          cooling.fraction.50 = 0.5,  # Cooling set so that we will reach smaller rw.sd at the end than the end of the unit3 search.
          rw.sd = epi_rw
        )
      }
    }
  )
}

@@

      for (j in 1:nrow(lee_epi_ll)) {
        # Get parameters of interest
        pf_params <- coef(lee1_epi_mif[[j]])

@@

# First, we need to simulate the epidemic model using the fitted parameters.
registerDoRNG(18599687)
foreach(
  i = 1:length(lee1_epi_mif),
  .combine = rbind
) %dopar% {
  guess <- coef(lee1_epi_mif)
  sims <- simulate(
    h1_epi,
    nsim = N_LEE_SIMS,
    format = 'data.frame',
    params = guess
  )  %>%
    dplyr::filter(week == max(week)) %>%
    dplyr::select(-.id, -week)

@@

end_start <- t(coef(lee1_epi_mif))

@@

# TODO: Only save parameter values and "likelihoods".
if (RUN_LEVEL != 3 & file.exists("model1/run_level_3/lee1_end_fit.rds")) {
  lee1_end_mif <- readRDS("model1/run_level_3/lee1_end_fit.rds")
} else {
  lee1_end_mif <- bake(
    file = paste0("model1/", rl_dir, "lee1_end_fit.rds"),
    {
      foreach(
        i = 1:nrow(end_start),
        .packages = c('pomp'),
        .combine = c
      ) %dopar% {

@@

        start_states %>%
          filter(param_set == i) %>%
          select(-param_set) %>%
          sapply(., median, na.rm = TRUE) -> rinit_parms

@@

        coef(h1_end) <- r_params
        mif2(
          h1_end,
          Np = NP_MIF,
          Nmif = NMIF,
          cooling.type = "hyperbolic",
          cooling.fraction.50 = 0.05,  # Cooling set so that we will reach smaller rw.sd at the end than the end of the unit3 search.
          rw.sd = end_rw
        )
        # output[['m2']] <- m2
        #
        # coef(h1_end) <- coef(m2)
        # sims <- simulate(  # Step 5
        #   h1_end, nsim = N_LEE_SIMS, format = 'data.frame',
        #   params = coef(m2)
        # )
        # sims$param_set <- i
        #
        # output[['sims']] <- sims
        # output
      }
    }
  )
}

rm(end_rw, start_states)
gc()

@@

      # Create data.frame to save results
      lee_end_ll <- data.frame(
        "pfLL" = rep(0, length(lee1_end_mif)),
        "pfse" = rep(0, length(lee1_end_mif))
      )

      # Perform particle filter to get likelihood estimates
      for (j in 1:nrow(lee_end_ll)) {
        # Get parameters of interest
        pf_params <- coef(lee1_end_mif[[j]])

@@

haiti3_fit <- bake(
  file = paste0("model3/", rl_dir, "haiti3_fit.rds"), {
    fit_haiti3(
      search1 = SEARCH1,
      search2 = SEARCH2,
      search3 = SEARCH3,
      search4 = SEARCH4,
      search5 = SEARCH5,
      n_searches = N_SEARCHES,
      ncores = cores
    )
  }
)

@@

<<Load Model Parameters, echo=FALSE, results='hide'>>=
# load('model3/output/PanelLocalAll_PF.rda')  # Final model 3 evaluations
# load('model3/output/PanelLocalAll.rda')  # Final model 3 parameters (MIF search)

@@

# Fit model 2. Note that this takes the same amount of time at all run
# levels.
h2_fit <- bake(
  file = paste0("model2/", rl_dir, "model2_fit.rds"), {
    fit_haiti2()
  },
  timing = FALSE
)

@@

<<Model 3 PanelSims,echo=FALSE>>=
h3_panel_sims <- switch(RUN_LEVEL, 20, 100, 500)

departements <-
  c(
    'Artibonite',
    'Centre',
    'Grande_Anse',
    'Nippes',
    'Nord',
    'Nord-Est',
    'Nord-Ouest',
    'Ouest',
    'Sud',
    'Sud-Est'
  )

registerDoRNG(5317865)

if (RUN_LEVEL != 3 & file.exists("model3/run_level_3/panelPompSims.rds")) {
  h3_panel_sims <- readRDS("model3/run_level_3/panelPompSims.rds")
} else {
  h3_panel_sims <- bake(
    file = paste0('model3/', rl_dir, 'panelPompSims.rds'), {
      foreach(dep = departements, .combine = rbind) %dopar% {
        SIRB <- unitobjects(h3)[[dep]]
        shared <- h3@shared
        specific <- h3@specific[, dep]

        pomp::simulate(
          SIRB,
          params = c(shared, specific),
          nsim = h3_panel_sims,
          format = 'data.frame'
        ) -> sims

        sims$dep <- dep
        sims
      }
    }
  )
}

h3_panel_quantiles <- h3_panel_sims %>%
  mutate(
    date = as.Date(
      lubridate::round_date(lubridate::date_decimal(time), unit = 'day')
    )
  ) %>%
  group_by(dep, date) %>%
  summarise(
    q05 = quantile(cases, 0.025, na.rm = T),
    mean = mean(cases, na.rm = T),
    q50 = quantile(cases, 0.5, na.rm = T),
    q95 = quantile(cases, 0.975, na.rm = T)
  ) %>%
  ungroup() %>%
  mutate(
    dep = gsub("-", "_", dep)
  ) %>%
  filter(date >= as.Date(
    lubridate::round_date(
      lubridate::date_decimal(h3@unit.objects$Artibonite@t0), unit = 'day'
    )
  )
  )

% \begin{figure}[ht]
% ggplot() +
%   geom_line(data = h3_panel_quantiles, aes(x = date, y = q50 + 1), col = 'blue') +
%   geom_line(data = dep_plot_df, aes(x = date, y = cases + 1), col = 'black') +
%   geom_ribbon(data = h3_panel_quantiles, aes(x = date, ymin = q05 + 1, ymax = q95 + 1),
%               alpha = 0.4) +
%   facet_wrap(~factor(dep, levels = plot_order), nrow = 2,
%              labeller = dep_labeller) +
%   ylab('Reported cholera cases') +
%   theme(axis.title.x = element_blank()) +
%   scale_x_date(date_labels = "'%y", breaks = seq.Date(from = as.Date("2011-01-01"), as.Date("2019-01-01"), by = '2 years')) +
%   scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))
%
% \caption{\label{h3panelsims}
% Simulations from initial conditions using panelPOMP version of Model~3.
% The black curve represents true case count, the blue curve the median of 500 simulations from the model, and the gray ribbons representing $95\%$ confidence interval.
% }
% \end{figure}

<<Remove Model3 panelSims, cache=FALSE, include=FALSE, message=FALSE, echo=FALSE>>=
rm(h3_panel_quantiles, h3_panel_sims)
gc()
@

Fitting Model~3 in this fashion simplifies the problem of parameter estimation, but it also introduces additional technicalities that must be addressed.
One concern is that of obtaining model forecasts, which was the primary goal of \citet{lee20}.
% "primary output" -> "primary goal", based on Anna's suggestion.
The simplified PanelPOMP version of Model~3 relies on the observed cholera cases as a covariate, which are unavailable for use in forecasts.
Similar to what was done in \cite{lee20}, we use the parameter estimates obtained with the independent departmental approximation of Model~3 as an estimate for the parameters in the fully coupled version of the model, which was implemented using the \code{spatPomp} package.
Simulations from the fully coupled version of the model using the estimated parameter vector $\tilde{\paramVec}$, are displayed in Fig.~\ref{h3spatsims}.

<<Model 3 SpatPOMP sims, echo=FALSE>>=
# load('output/FinalPanelAsSpatPomp_PF.rda')

h3_spat_nsim <- switch(RUN_LEVEL, 20, 100, 500)

@@

mod2_epi_ll <- -epi_ofun(par = h2_fit$h2_params)
# mod2_end_ll <- -end_ofun(par = h2_fit$end_params)

mod2_ll <- mod2_epi_ll - sum(log(h2_epi@data + 1), na.rm = TRUE)

# Evaluate and sum objective functions to get total model likelihood
# mod2_ll <- -(epi_ofun(par = h2_fit$epi_params) + end_ofun(par = h2_fit$end_params))

@@

# Load pre-calculated model 3 likelihood
# mod3_ll <- ll_mod3 %>%
#   arrange(-logLik) %>%
#   slice_head(n = 1) %>%
#   pull(logLik)

@@

# load('model2/output/mod2_leeFit.rda')
mod2_lee_aic <- 2 * h2_fit$leeFit_n_params - 2 * h2_fit$leeFit_ll
# mod2_lee_aic <- 2 * mod2_leeFit_num_parms - 2 * mod2_leeFit_ll
@

>>>>

# Check if RUN_LEVEL == 3 results exists, if so, use them.
# Include a flag to check if RUN_LEVEL == 3. If so, we might want to recompute.
# TODO: Only save parameters and likelihoods, not entire MIF objects.
if (RUN_LEVEL != 3 & file.exists("model1/run_level_3/lee1_epi_fit.rds")) {
  lee1_epi_mif <- readRDS("model1/run_level_3/lee1_epi_fit.rds")
} else {
  lee1_epi_mif <- bake(
    file = paste0("model1/", rl_dir, "lee1_epi_fit.rds"),
    {
      lee1_epi_mif_results <- list()

      foreach(
        i = 1:nrow(epi_params),
        .packages = c('pomp'),
        .combine = c
      ) %dopar% {
        r_params <- unlist(epi_params[i, ])
        coef(h1_epi) <- r_params
        mif2(
          h1_epi,
          Np = NP_MIF,
          Nmif = NMIF,
          cooling.fraction.50 = 0.5,  # Cooling set so that we will reach smaller rw.sd at the end than the end of the unit3 search.
          rw.sd = epi_rw
        )
      } -> lee1_epi_mif_out

      foreach(i = 1:length(lee1_epi_mif_out), .combine = rbind) %do% {
        guess <- coef(lee1_epi_mif_out[[i]])
        sims <- simulate(
          h1_epi,
          n = N_LEE_SIMS,
          format = 'data.frame',
          params = guess
        )  %>%
          dplyr::filter(week == max(week)) %>%
          dplyr::select(-.id, -week)

      lee1_epi_mif_results$lee1_epi_mif_params <- t(coef(lee1_epi_mif_out))
      lee1_epi_mif_results$start_states <- start_states
      lee1_epi_mif_results
    }
  )
}

@@

      for (j in 1:nrow(lee1_epi_mif$lee1_epi_mif_params)) {
        # Get parameters of interest
        pf_params <- lee1_epi_mif$lee1_epi_mif_params[j, ]

@@

@@

end_start <- lee1_epi_mif$lee1_epi_mif_params

@@

# TODO: Only save parameter values and "likelihoods".
if (RUN_LEVEL != 3 & file.exists("model1/run_level_3/lee1_end_fit.rds")) {
  lee1_end_mif <- readRDS("model1/run_level_3/lee1_end_fit.rds")
} else {
  lee1_end_mif <- bake(
    file = paste0("model1/", rl_dir, "lee1_end_fit.rds"),
    {
      foreach(
        i = 1:nrow(end_start),
        .packages = c('pomp'),
        .combine = rbind
      ) %dopar% {

@@

        lee1_epi_mif$start_states %>%
          filter(param_set == i) %>%
          select(-param_set) %>%
          sapply(., median, na.rm = TRUE) -> rinit_parms

@@

        coef(h1_end) <- r_params
        mf_end <- mif2(
          h1_end,
          Np = NP_MIF,
          Nmif = NMIF,
          cooling.type = "hyperbolic",
          cooling.fraction.50 = 0.05,  # Cooling set so that we will reach smaller rw.sd at the end than the end of the unit3 search.
          rw.sd = end_rw
        )

        coef(mf_end)
      }
    }
  )
}

rm(end_rw)
gc()

@@

      # Create data.frame to save results
      lee_end_ll <- data.frame(
        "pfLL" = rep(0, nrow(lee1_end_mif)),
        "pfse" = rep(0, nrow(lee1_end_mif))
      )

      # Perform particle filter to get likelihood estimates
      for (j in 1:nrow(lee_end_ll)) {
        # Get parameters of interest
        pf_params <- lee1_end_mif[j, ]

@@

if (RUN_LEVEL != 3 & file.exists("model3/run_level_3/haiti3_fit.rds")) {
  haiti3_fit <- readRDS("model3/run_level_3/haiti3_fit.rds")
} else {
  haiti3_fit <- bake(
    file = paste0("model3/", rl_dir, "haiti3_fit.rds"), {
      fit_haiti3(
        search1 = SEARCH1,
        search2 = SEARCH2,
        search3 = SEARCH3,
        search4 = SEARCH4,
        search5 = SEARCH5,
        n_searches = N_SEARCHES,
        ncores = cores
      )
    }
  )
}

@@

<<Load Model Parameters, echo=FALSE, results='hide'>>=

@@

# Fit model 2. Note that this takes the same amount of time at all run
# levels.
h2_fit <- bake(
  file = paste0("model2/model2_fit.rds"), {
    fit_haiti2()
  },
  timing = FALSE
)

@@

Fitting Model~3 in this fashion simplifies the problem of parameter estimation, but it also introduces additional technicalities that must be addressed.
One concern is that of obtaining model forecasts, which was the primary goal of \citet{lee20}.
The simplified PanelPOMP version of Model~3 relies on the observed cholera cases as a covariate, which are unavailable for use in forecasts.
Similar to what was done in \cite{lee20}, we use the parameter estimates obtained with the independent departmental approximation of Model~3 as an estimate for the parameters in the fully coupled version of the model, which was implemented using the \code{spatPomp} package.
Simulations from the fully coupled version of the model using the estimated parameter vector $\tilde{\paramVec}$, are displayed in Fig.~\ref{h3spatsims}.

<<Model 3 SpatPOMP sims, echo=FALSE>>=
h3_spat_nsim <- switch(RUN_LEVEL, 20, 100, 500)

@@

mod2_epi_ll <- -epi_ofun(par = h2_fit$h2_params)
mod2_ll <- mod2_epi_ll - sum(log(h2_epi@data + 1), na.rm = TRUE)

@@

@@

mod2_lee_aic <- 2 * h2_fit$leeFit_n_params - 2 * h2_fit$leeFit_ll
@
